---
title: "Week7AssignmentCode"
author: "Jonnathan Zuna Largo"
format: html
editor: visual
---

## Introduction

My goal is to create an HTML and also a JSON file by inputting it manually the same structure book data. After creating the files, I will use R to load each file into a dataframe and ensure that both files give me identical results. The requirement for this projects are 3 books in which my choices are: "Astrophysics for people in a hurry" "Atomic Habits"  and "An Introduction to Statistical Learning." I will search the authors, publication year, publisher, and ISBN for each one and then work for the process of creating files and then comparing the results.  

## Body

### Files

In this section I'll create the html file and the json file by using the information from each book I selected

```{html}
<!DOCTYPE html>
<html>
  <head>
    <title>Books</title>
  </head>
  <body>
    <table>
      <tr>
        <th>Title</th>
        <th>Authors</th>
        <th>Year</th>
        <th>Publisher</th>
        <th>ISBN</th>
      </tr>
      <tr>
        <td>Astrophysics for people in a hurry</td>
        <td>Neil deGrasse Tyson</td>
        <td>2017</td>
        <td>W. W. Norton & Company</td>
        <td>978-0393609394</td>
      </tr>
      <tr>
        <td>Atomic Habits</td>
        <td>James Clear</td>
        <td>2018</td>
        <td>Random House Business Books</td>
        <td>978-1847941848</td>
      </tr>
      <tr>
        <td>An Introduction to Statistical Learning</td>
        <td>Gareth James, Daniela Witten, Trevor Hastie, Robert Tibshirani</td>
        <td>2021</td>
        <td>Springer</td>
        <td>978-1071614174</td>
      </tr>
    </table>
  </body>
</html>

```


```{json}
[
  {
    "Title": "Astrophysics for people in a hurry",
    "Authors": "Neil deGrasse Tyson",
    "Year": 2017,
    "Publisher": "W. W. Norton & Company",
    "ISBN": "978-0393609394"
  },
  {
    "Title": "Atomic Habits",
    "Authors": "James Clear",
    "Year": 2018,
    "Publisher": "Random House Business Books",
    "ISBN": "978-1847941848"
  },
  {
    "Title": "An Introduction to Statistical Learning",
    "Authors": "Gareth James, Daniela Witten, Trevor Hastie, Robert Tibshirani",
    "Year": 2021,
    "Publisher": "Springer",
    "ISBN": "978-1071614174"
  }
]
```

### Code 

For the following section the packages rvest will be used for html. The function read_html() parses the file and html_table() extracts the table it finds.

```{r}
library(rvest)
html_raw <- read_html("BookInformation.html")
```


```{r}
df_html <- html_table(html_raw)[[1]]
head(df_html)
```

The the json file the package jsonlite handles it perfectly. The function fromJSON() reads the file.

```{r}
library(jsonlite)
df_json <- fromJSON("BookInformation.json")

head(df_json)
```

First, I want to compare the structure for both data frames side by side

```{r}
names(df_html)
names(df_json)
```

Both structures are correct so now I can compare data types

```{r}
class(df_html)
class(df_json)
```

As shown above df_html is coming in a different format than the json file, to compare them both I'll change the html to the same format as the json file

```{r}
df_html <- as.data.frame(df_html)
```

Now that both tables are normalized and in the same format, I'll compare them both and both should be identical

```{r}
all.equal(df_html, df_json)
```

```{r}
identical(df_html, df_json)
```

## Conclusion

Working in this exercise game a more clear understanding on how the same data can live in two different formats and what takes to make them both comparable in a consistent format. Loading the files were straight forward however the comparison part showed a subtle structural difference that I wouldn't have realized if I did not inspect the data frame classes directly. Moving forward, being aware of type and class consistency when loading from multiple sources will be a crucial step to build before working into web scraping and API projects. 