---
title: "Week9AssignmentCode"
author: "Jonnathan Zuna Largo"
format: html
editor: visual
---

## Introduction

This assignment require us to select one of the NYT endpoints, authenticate securely using an API key, make a live request from R, parse the JSON response, and deliver a clean, tidy data frame ready for analysis. The goal is not just to pull data but to ask a meaningful question and use the API response to begin answering it. Rather than a simple retrieval exercise, this is an end to end data pipeline task that mirrors what real data work looks like when consuming third party web services.

## Code

### Libraries

```{r}
library(httr)
library(jsonlite)
library(dplyr)
library(tibble)
library(ggplot2)
```

### API Key

This section we make a

```{r}
api_key <- Sys.getenv("NYT_API_KEY")

response <- GET(
  url = "https://api.nytimes.com/svc/books/v3/lists/current/hardcover-fiction.json",
  query = list(`api-key` = api_key)
)

status_code(response)
```

Parsing the Json response and obseving what was parsed

```{r}
raw_content <- content(response, as = "text", encoding = "UTF-8")
parsed <- fromJSON(raw_content, flatten = TRUE)
names(parsed)
```

Inside the results I can see that my target "books" is there

```{r}
names(parsed$results)
```

Now we can observe the fileds available in books

```{r}
names(parsed$results$books)
```

Only selected fileds will be used are stated in my introduction. This section creates the datafram books_df

```{r}
books_df <- parsed$results$books |>
  select(title, author, publisher, rank, rank_last_week, weeks_on_list, description) |>
  as_tibble() |>
  mutate(
    rank = as.integer(rank),
    rank_last_week = as.integer(rank_last_week),
    weeks_on_list = as.integer(weeks_on_list)
  )

glimpse(books_df)
```

```{r}
author_summary <- books_df |>
  arrange(desc(weeks_on_list)) |>
  select(title, author, rank, weeks_on_list)

print(author_summary)
```

To make the graph it is helpful to sorted and maake the bars horizontal for better so it can be readable

```{r}
books_df |>
  arrange(desc(weeks_on_list)) |>
  ggplot(aes(x = reorder(title, weeks_on_list), y = weeks_on_list, fill = weeks_on_list)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "steelblue", high = "darkblue") +
  labs(
    title = "Weeks on the NYT Hardcover Fiction Bestseller List",
    x = NULL,
    y = "Weeks on List",
    fill = "Weeks",
    caption = paste("Published date:", parsed$results$published_date)
  ) +
  theme_minimal()
```

## Conclusion

The data returned by the NYT Books API was largely clean and required minimal intervention. The response JSON is nested, with the books data living inside parsed-results-books, which required navigating two levels to parse the object before extracting it. I seleted 7 fileds total from the available fileds and those were: title, author, publisher, rank, rank_last_week, weeks_on_list, and description. The three numeric fields, rank, rank_last_week, and weeks_on_list, I changed them to integer using as.integer() since JSON parsing via jsonlite returns them as numeric by default. A rank_last_week value of zero indicates the book is newly entered on the list this week and did not hold a position the prior week, which is meaningful context rather than a missing value, so no imputation was applied.
